---
title: "hw3"
output: html_document
---

5. It is well-known that ridge regression tends to give similar coefficient
values to correlated variables, whereas the lasso may give quite different coefficient values to correlated variables. We will now explore this property in a very simple setting.
Suppose that n = 2, p = 2, x11 = x12, x21 = x22. Furthermore,
suppose that y1 +y2 = 0 and x11 +x21 = 0 and x12 +x22 = 0, so that
the estimate for the intercept in a least squares, ridge regression, or
lasso model is zero: βˆ0 = 0.

(a) Write out the ridge regression optimization problem in this setting.
(b) Argue that in this setting, the ridge coefficient estimates satisfy
βˆ1 = βˆ2.
(c) Write out the lasso optimization problem in this setting.
(d) Argue that in this setting, the lasso coefficients βˆ1 and βˆ2 are
not unique—in other words, there are many possible solutions
to the optimization problem in (c). Describe these solutions.


8. In this exercise, we will generate simulated data, and will then use
this data to perform best subset selection.

(a) Use the rnorm() function to generate a predictor X of length
n = 100, as well as a noise vector ϵ of length n = 100.

```{r}
X <- rnorm(100)
epsilon <- rnorm(100)
```


(b) Generate a response vector Y of length n = 100 according to
the model Y = β0 + β1X + β2X2 + β3X3 + ϵ, where β0, β1, β2, and β3 are constants of your choice.

```{r}
b_0 <- 5
b_1 <- 3
b_2 <- 6
b_3 <- 7
y <- b_0 + b_1*X + b_2*X^2 + b_3*X^3 + epsilon
```


(c) Use the regsubsets() function to perform best subset selection
in order to choose the best model containing the predictors
X, X2,...,X10. What is the best model obtained according to
Cp, BIC, and adjusted R2? Show some plots to provide evidence
for your answer, and report the coefficients of the best model obtained. Note you will need to use the data.frame() function to create a single data set containing both X and Y .

```{r}
library(leaps)

set.seed(42)

df <- data.frame(y=y, x=X)
subset_fit <- regsubsets(y ~ I(X) + I(X^2)+ I(X^3) + I(X^4) + I(X^5) + I(X^6) + I(X^7) + I(X^8) + I(X^9) + I(X^10), data=df, nvmax=10)
reg_summary <- summary(subset_fit)

# Best model according to cp, bic, and adjr2
reg_cp <- which.min(reg_summary$cp)
reg_bic <- which.min(reg_summary$bic)
reg_adjr2 <- which.max(reg_summary$adjr2)

reg_cp
reg_bic
reg_adjr2

# par(mfrow=c(2,2))

plot(subset_fit)

plot(reg_summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
points(which.max(reg_summary$adjr2),reg_summary$adjr2[which.max(reg_summary$adjr2)], col="red",cex=2,pch=20)

plot(reg_summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
points(which.min(reg_summary$cp),reg_summary$cp[which.min(reg_summary$cp)],col="red",cex=2,pch=20)

plot(reg_summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
points(which.min(reg_summary$bic),reg_summary$bic[which.min(reg_summary$bic)],col="red",cex=2,pch=20)


# coefficients for best 3 variable model 
coef(subset_fit, 3)
```


(d) Repeat (c), using forward stepwise selection and also using backwards stepwise selection. How does your answer compare to the
results in (c)?

Forward selection:
```{r}
forward_fit <- regsubsets(y ~ I(X) + I(X^2)+ I(X^3) + I(X^4) + I(X^5) + I(X^6) + I(X^7) + I(X^8) + I(X^9) + I(X^10), data=df, nvmax=10, method="forward")
forward_summary <- summary(forward_fit)

# Best model according to cp, bic, and adjr2
which.min(forward_summary$cp)
which.min(forward_summary$bic)
which.max(forward_summary$adjr2)

# coefficients for best 3 variable model 
coef(forward_fit, 3)
```

Backward selection:
```{r}
backward_fit <- regsubsets(y ~ I(X) + I(X^2)+ I(X^3) + I(X^4) + I(X^5) + I(X^6) + I(X^7) + I(X^8) + I(X^9) + I(X^10), data=df, nvmax=10, method="backward")
backward_summary <- summary(backward_fit)

# Best model according to cp, bic, and adjr2
which.min(backward_summary$cp)
which.min(backward_summary$bic)
which.max(backward_summary$adjr2)


# coefficients for best 3 variable model 
coef(backward_fit, 3)
```

The performance of each model tells us that a 3 variable model was the best model. 
The regular use of `rsubsets()` (neither forward nor backward stepwise) and the forward
stepwise says that X, X^2, X^3, X^5, and X^10 were the best variables whereas the
backwards stepwise says that X, X^2, X^3, X^5, and X^7 were the best variables.

(e) Now fit a lasso model to the simulated data, again using X, X2,
...,X10 as predictors. Use cross-validation to select the optimal
value of λ. Create plots of the cross-validation error as a function
of λ. Report the resulting coefficient estimates, and discuss the
results obtained.

```{r}
library(glmnet)

X <- model.matrix(y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10), data=df)[, -1]
y <- df$y

set.seed(42)
cv_lasso <- cv.glmnet(X,y,alpha=1)
plot(cv_lasso, sign.lambda=1)
```
```{r}
min.lambda <- cv_lasso$lambda.min
min.lambda
```


```{r}
prediction <- predict(cv_lasso, s = min.lambda, type = "coefficients")[1:11, ]
prediction
```

The the model that uses lambda min only contains 5 out of the ten total predictors used: x, x^2, x^3, x^4, x^5. The Lasso
method being used gives us information of what predictors are not useful by making them all 0.

(f) Now generate a response vector Y according to the model
Y = β0 + β7X7 + ϵ,
and perform best subset selection and the lasso. Discuss the
results obtained.

```{r}
set.seed(42)
X <- rnorm(100)
e <- rnorm(100)

b_0 <- 5
b_7 <- 3
y <- b_0 + b_7*X^7 + e
```

Best subset selection:
```{r}
df <- data.frame(y=y, x=X)
diff_fit <- regsubsets(y ~ I(X) + I(X^2)+ I(X^3) + I(X^4) + I(X^5) + I(X^6) + I(X^7) + I(X^8) + I(X^9) + I(X^10), data=df, nvmax=10)
diff_summary <- summary(diff_fit)

# Best model according to cp, bic, and adjr2
which.min(diff_summary$cp)
which.min(diff_summary$bic)
which.max(diff_summary$adjr2)

plot(diff_summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
points(which.max(diff_summary$adjr2),diff_summary$adjr2[which.max(diff_summary$adjr2)], col="red",cex=2,pch=20)

plot(diff_summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
points(which.min(diff_summary$cp),diff_summary$cp[which.min(diff_summary$cp)],col="red",cex=2,pch=20)

plot(diff_summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
points(which.min(diff_summary$bic),diff_summary$bic[which.min(diff_summary$bic)],col="red",cex=2,pch=20)


# coefficients for best 4 variable model 
coef(diff_fit, 4)
```


Lasso:
```{r}
X <- model.matrix(y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10), data=df)[, -1]
y <- df$y

set.seed(42)
diff_lasso <- cv.glmnet(X,y,alpha=1)
plot(diff_lasso, sign.lambda=1)
```
```{r}
min.lambda <- diff_lasso$lambda.min
min.lambda
```

```{r}
prediction <- predict(diff_lasso, s = min.lambda, type = "coefficients")[1:11, ]
prediction
```

In this instance, when the min lambda was used it only chose one predictor, x^7. This is because the response vector created in this problem contained x^7.

9. In this exercise, we will predict the number of applications received
using the other variables in the College data set.

(a) Split the data set into a training set and a test set.
```{r}
require(ISLR)
data(College)
attach(College)

set.seed(42)

college_train = sample(1:nrow(College), nrow(College) * 0.7)
college_test = -college_train
y_test <- Apps[college_test]
```


(b) Fit a linear model using least squares on the training set, and
report the test error obtained.
```{r}
college_ols <- lm(Apps ~ ., data = College, subset = college_train)

summary(college_ols)
```
```{r}
college_predict <- predict(college_ols, College[college_test, ])
college_error <- mean((college_predict - y_test)^2)
college_error
```
The test MSE is 1583501

(c) Fit a ridge regression model on the training set, with λ chosen
by cross-validation. Report the test error obtained.

```{r}
matrix <- model.matrix(Apps~., data=College)[, -1]
matrix.train <- matrix[college_train,]
y <- College$Apps
y_train <- y[college_train]

ridge <- cv.glmnet(matrix.train, y_train, alpha=0)
plot(ridge, sign.lambda = 1)
```

```{r}
ridge.lambda <- ridge$lambda.min

ridge_predict <- predict(ridge, s = ridge.lambda, newx = matrix[college_test, ])
ridge_error <- mean((ridge_predict - y_test)^2)
ridge_error
```

The test error for ridge is 2828195. 

(d) Fit a lasso model on the training set, with λ chosen by crossvalidation. Report the test error obtained, along with the number of non-zero coefficient estimates.

```{r}
lasso <- cv.glmnet(matrix.train, y_train, alpha=1)
plot(lasso, sign.lambda=1)
```
```{r}
lasso.lambda <- lasso$lambda.min
lasso.lambda
```
The min lambda for lasso is 9.660522

```{r}
lasso_predict <- predict(lasso, s = lasso.lambda, newx = matrix[college_test, ])
lasso_error <- mean((lasso_predict - y_test)^2)
lasso_error
```

The test MSE for lasso is 1675404

```{r}
lasso.c <- predict(lasso,type="coefficients", s=lasso.lambda)[1:17,]
length(lasso.c[lasso.c != 0])
```
Lasso produced 15 coefficient estimates that are not 0

(e) Fit a PCR model on the training set, with M chosen by crossvalidation. Report the test error obtained, along with the value
of M selected by cross-validation.

```{r}
library(pls)

pcr_fit <- pcr(Apps~., data=College, subset=college_train, scale=TRUE, validation ="CV")
summary(pcr_fit)
```
```{r}
validationplot(pcr_fit,val.type="MSEP")
```
The lowest CV error is when all 17 variables are used

```{r}
pcr_pred <- predict(pcr_fit,matrix[college_test,], ncomp=17) 
pcr_error <- mean((pcr_pred-y_test)^2)
pcr_error
```
The test MSE is 1583501

(f) Fit a PLS model on the training set, with M chosen by crossvalidation. Report the test error obtained, along with the value
of M selected by cross-validation.

```{r}
pls_fit <- plsr(Apps~., data=College, subset=college_train, scale=TRUE, validation="CV")
summary(pls_fit)
```
```{r}
validationplot(pls_fit,val.type="MSEP")
```
```{r}
pls_pred <- predict(pls_fit,matrix[college_test,],ncomp=10) 
pls_error <- mean((pls_pred-y_test)^2)
pls_error
```


(g) Comment on the results obtained. How accurately can we predict the number of college applications received? Is there much
difference among the test errors resulting from these five approaches?

```{r}
results <- c(college_error, ridge_error, lasso_error, pcr_error, pls_error)
names(results) <- c("lm", "ridge", "lasso", "pcr", "pls")
barplot(results)
```
```{r}
print(sort(results))
```

Ridge ended up performing the worst by having the highest error whereas the other
four methods were within a small range of each other. PCR performed the best with having the lowest 
test error.
