---
title: "ISLR Chapter 7 Lab: Moving Beyond Linearity"
author: "Hsin-Hsiung Huang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300,
  fig.align = "center"
)

required_pkgs <- c("ggplot2", "splines", "mgcv", "knitr")
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(sprintf("Please install package '%s' before knitting this file.", pkg))
  }
}

if (requireNamespace("ISLR", quietly = TRUE)) {
  data("Wage", package = "ISLR")
} else if (requireNamespace("ISLR2", quietly = TRUE)) {
  data("Wage", package = "ISLR2")
} else {
  stop("Please install either the ISLR or ISLR2 package to load the Wage data.")
}

library(ggplot2)
library(splines)
library(mgcv)
library(knitr)

out_dir <- "chapter7_outputs"
fig_dir <- file.path(out_dir, "figures")
tbl_dir <- file.path(out_dir, "tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tbl_dir, recursive = TRUE, showWarnings = FALSE)

save_plot <- function(p, name, width = 7, height = 4.5) {
  ggsave(filename = file.path(fig_dir, paste0(name, ".pdf")), plot = p,
         width = width, height = height, units = "in")
  ggsave(filename = file.path(fig_dir, paste0(name, ".png")), plot = p,
         width = width, height = height, units = "in", dpi = 300)
  invisible(p)
}

save_table <- function(x, name, digits = 3, align = NULL) {
  tex <- knitr::kable(x, format = "latex", booktabs = TRUE,
                      digits = digits, align = align)
  writeLines(tex, file.path(tbl_dir, paste0(name, ".tex")))
  invisible(x)
}

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2, na.rm = TRUE))
mae <- function(y, yhat) mean(abs(y - yhat), na.rm = TRUE)
rsq <- function(y, yhat) {
  ok <- is.finite(yhat)
  y <- y[ok]
  yhat <- yhat[ok]
  1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
}
logloss <- function(y, p) {
  p <- pmin(pmax(p, 1e-6), 1 - 1e-6)
  -mean(y * log(p) + (1 - y) * log(1 - p))
}
brier <- function(y, p) mean((y - p)^2)
accuracy <- function(y, p, threshold = 0.5) mean((p >= threshold) == (y == 1))

predict_loess_safe <- function(model, newdata, train_x = NULL) {
  pred <- predict(model, newdata = newdata)
  if (all(!is.na(pred))) return(pred)
  if (is.null(train_x)) stop("train_x must be supplied when loess predictions contain NA values.")
  fitted_train <- predict(model, newdata = data.frame(age = train_x))
  approx(x = train_x, y = fitted_train, xout = newdata$age, rule = 2)$y
}

make_step_model <- function(train_data, bins) {
  br <- seq(min(train_data$age), max(train_data$age), length.out = bins + 1)
  br[1] <- min(Wage$age) - 1e-6
  br[length(br)] <- max(Wage$age) + 1e-6
  train_data$age_bin <- cut(train_data$age, breaks = br, include.lowest = TRUE)
  fit <- lm(wage ~ age_bin, data = train_data)
  list(fit = fit, breaks = br)
}

predict_step_model <- function(obj, newdata) {
  nd <- newdata
  nd$age_bin <- cut(nd$age, breaks = obj$breaks, include.lowest = TRUE)
  predict(obj$fit, newdata = nd)
}

cv_assignments <- function(n, K = 10, seed = 5703) {
  set.seed(seed)
  sample(rep(1:K, length.out = n))
}

kfold_rmse <- function(model_type, values, data = Wage, K = 10) {
  folds <- cv_assignments(nrow(data), K = K, seed = 5703)
  out <- data.frame(setting = values, cv_rmse = NA_real_)

  for (i in seq_along(values)) {
    val <- values[i]
    fold_err <- numeric(K)

    for (k in seq_len(K)) {
      train <- data[folds != k, ]
      valid <- data[folds == k, ]

      if (model_type == "poly") {
        fit <- lm(wage ~ poly(age, val), data = train)
        pred <- predict(fit, newdata = valid)
      } else if (model_type == "step") {
        fit <- make_step_model(train, bins = val)
        pred <- predict_step_model(fit, valid)
      } else if (model_type == "ns") {
        fit <- lm(wage ~ ns(age, df = val), data = train)
        pred <- predict(fit, newdata = valid)
      } else if (model_type == "ss") {
        fit <- smooth.spline(train$age, train$wage, df = val)
        pred <- predict(fit, x = valid$age)$y
      } else if (model_type == "loess") {
        fit <- loess(wage ~ age, data = train, span = val, degree = 1)
        pred <- predict_loess_safe(fit, data.frame(age = valid$age), train_x = train$age)
      } else {
        stop("Unknown model_type")
      }

      fold_err[k] <- rmse(valid$wage, pred)
    }

    out$cv_rmse[i] <- mean(fold_err)
  }

  out
}

Wage$high_wage <- as.integer(Wage$wage > 250)
set.seed(5703)
train_id <- sample.int(nrow(Wage), floor(0.80 * nrow(Wage)))
train <- Wage[train_id, ]
test  <- Wage[-train_id, ]

age_grid <- data.frame(age = seq(min(Wage$age), max(Wage$age), length.out = 200))
age_profile <- data.frame(
  age = seq(min(Wage$age), max(Wage$age), length.out = 200),
  year = round(median(Wage$year)),
  education = levels(Wage$education)[1]
)
```

# Overview

These notes implement the main Chapter 7 nonlinear methods using the `Wage` data. The file is designed for two purposes:

1. to serve as a fully runnable lab for students, and
2. to generate figures and LaTeX tables for `Chapter7.tex`.

When you knit this document, the code writes figures to `chapter7_outputs/figures/` and LaTeX tables to `chapter7_outputs/tables/`.

## Data overview

```{r data-summary}
overview_tbl <- data.frame(
  variable = c("n", "age mean", "age sd", "wage mean", "wage sd", "high wage rate"),
  value = c(
    nrow(Wage),
    mean(Wage$age),
    sd(Wage$age),
    mean(Wage$wage),
    sd(Wage$wage),
    mean(Wage$high_wage)
  )
)
save_table(overview_tbl, "tbl_overview_data_summary", digits = 3)
kable(overview_tbl, digits = 3)
```

```{r raw-wage-plot}
p_raw <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15) +
  labs(title = "Raw Wage data", x = "Age", y = "Wage")
save_plot(p_raw, "fig_overview_raw_wage")
p_raw
```

# Polynomial regression

## Degree tuning

```{r poly-degree-cv}
poly_cv <- kfold_rmse("poly", values = 1:8, data = Wage, K = 10)
colnames(poly_cv) <- c("degree", "cv_rmse")
save_table(poly_cv, "tbl_poly_degree_cv", digits = 3)

p_poly_cv <- ggplot(poly_cv, aes(degree, cv_rmse)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:8) +
  labs(title = "Polynomial degree tuning by 10-fold CV",
       x = "Polynomial degree", y = "CV RMSE")
save_plot(p_poly_cv, "fig_poly_degree_cv")

poly_cv
```

```{r poly-fit}
fit_poly4 <- lm(wage ~ poly(age, 4), data = Wage)
pred_poly4 <- predict(fit_poly4, newdata = age_grid, se = TRUE)

poly_df <- transform(
  age_grid,
  fit = pred_poly4$fit,
  lwr = pred_poly4$fit - 1.96 * pred_poly4$se.fit,
  upr = pred_poly4$fit + 1.96 * pred_poly4$se.fit
)

p_poly_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15) +
  geom_ribbon(data = poly_df,
              aes(x = age, ymin = lwr, ymax = upr),
              inherit.aes = FALSE,
              alpha = 0.20) +
  geom_line(data = poly_df,
            aes(x = age, y = fit),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Quartic polynomial regression", x = "Age", y = "Wage")
save_plot(p_poly_fit, "fig_poly_fit")
p_poly_fit
```

## Orthogonal versus raw polynomial basis

```{r raw-vs-orth}
fit_poly_raw  <- lm(wage ~ poly(age, 4, raw = TRUE), data = Wage)
fit_poly_orth <- lm(wage ~ poly(age, 4), data = Wage)
poly_basis_tbl <- data.frame(
  quantity = c("max abs difference in fitted values"),
  value = c(max(abs(predict(fit_poly_raw) - predict(fit_poly_orth))))
)
save_table(poly_basis_tbl, "tbl_poly_basis_compare", digits = 10)
kable(poly_basis_tbl, digits = 10)
```

## Logistic polynomial regression

```{r poly-logit}
fit_poly_logit <- glm(high_wage ~ poly(age, 3), data = Wage, family = binomial)
pred_poly_logit <- predict(fit_poly_logit, newdata = age_grid, se.fit = TRUE, type = "link")

poly_logit_df <- transform(
  age_grid,
  prob = plogis(pred_poly_logit$fit),
  lwr = plogis(pred_poly_logit$fit - 1.96 * pred_poly_logit$se.fit),
  upr = plogis(pred_poly_logit$fit + 1.96 * pred_poly_logit$se.fit)
)

p_poly_logit <- ggplot(poly_logit_df, aes(age, prob)) +
  geom_ribbon(aes(x = age, ymin = lwr, ymax = upr), alpha = 0.20) +
  geom_line(aes(x = age, y = prob), linewidth = 1) +
  labs(title = "Polynomial logistic regression", x = "Age", y = "P(wage > 250 | age)")
save_plot(p_poly_logit, "fig_poly_logit")
p_poly_logit
```

```{r poly-logit-eval}
poly_logit_prob_test <- predict(fit_poly_logit, newdata = test, type = "response")
poly_logit_eval <- data.frame(
  metric = c("log loss", "brier score", "accuracy"),
  value = c(
    logloss(test$high_wage, poly_logit_prob_test),
    brier(test$high_wage, poly_logit_prob_test),
    accuracy(test$high_wage, poly_logit_prob_test)
  )
)
save_table(poly_logit_eval, "tbl_poly_logit_eval", digits = 4)
kable(poly_logit_eval, digits = 4)
```

# Step functions

## Bin tuning

```{r step-tuning}
step_cv <- kfold_rmse("step", values = 3:8, data = Wage, K = 10)
colnames(step_cv) <- c("bins", "cv_rmse")
save_table(step_cv, "tbl_step_bins_cv", digits = 3)

p_step_cv <- ggplot(step_cv, aes(bins, cv_rmse)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 3:8) +
  labs(title = "Step-function tuning by 10-fold CV",
       x = "Number of bins", y = "CV RMSE")
save_plot(p_step_cv, "fig_step_bins_cv")

step_cv
```

## Four-bin step function fit

```{r step-fit}
step4_obj <- make_step_model(Wage, bins = 4)
step_df <- transform(age_grid, fit = predict_step_model(step4_obj, age_grid))

p_step_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15) +
  geom_line(data = step_df,
            aes(x = age, y = fit),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Step function with four bins", x = "Age", y = "Wage")
save_plot(p_step_fit, "fig_step_fit")
p_step_fit
```

# Regression splines and natural splines

## Natural spline tuning by degrees of freedom

```{r ns-tuning}
ns_cv <- kfold_rmse("ns", values = 3:10, data = Wage, K = 10)
colnames(ns_cv) <- c("df", "cv_rmse")
save_table(ns_cv, "tbl_ns_df_cv", digits = 3)

p_ns_cv <- ggplot(ns_cv, aes(df, cv_rmse)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 3:10) +
  labs(title = "Natural spline tuning by 10-fold CV",
       x = "Degrees of freedom", y = "CV RMSE")
save_plot(p_ns_cv, "fig_ns_df_cv")

ns_cv
```

## Cubic regression spline with explicit knots

```{r bs-fit}
fit_bs <- lm(wage ~ bs(age, knots = c(30, 40, 50, 60)), data = Wage)
bs_df <- transform(age_grid, fit = predict(fit_bs, newdata = age_grid))

p_bs_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15) +
  geom_line(data = bs_df,
            aes(x = age, y = fit),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Cubic regression spline", x = "Age", y = "Wage")
save_plot(p_bs_fit, "fig_bs_fit")
p_bs_fit
```

## Natural spline fit

```{r ns-fit}
fit_ns5 <- lm(wage ~ ns(age, df = 5), data = Wage)
ns_df <- transform(age_grid, fit = predict(fit_ns5, newdata = age_grid))

p_ns_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.15) +
  geom_line(data = ns_df,
            aes(x = age, y = fit),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Natural spline with 5 degrees of freedom", x = "Age", y = "Wage")
save_plot(p_ns_fit, "fig_ns_fit")
p_ns_fit
```

## Side-by-side comparison of age smoothers

```{r spline-compare}
fit_poly3 <- lm(wage ~ poly(age, 3), data = Wage)
compare_df <- age_grid
compare_df$poly3 <- predict(fit_poly3, newdata = age_grid)
compare_df$step4 <- predict_step_model(step4_obj, age_grid)
compare_df$ns5   <- predict(fit_ns5, newdata = age_grid)

p_compare <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.12) +
  geom_line(data = compare_df,
            aes(x = age, y = poly3, linetype = "Polynomial"),
            inherit.aes = FALSE) +
  geom_line(data = compare_df,
            aes(x = age, y = step4, linetype = "Step"),
            inherit.aes = FALSE) +
  geom_line(data = compare_df,
            aes(x = age, y = ns5, linetype = "Natural spline"),
            inherit.aes = FALSE) +
  labs(title = "Comparing nonlinear fits", x = "Age", y = "Wage", linetype = "Model")
save_plot(p_compare, "fig_spline_compare")
p_compare
```

# Smoothing splines

## Effective degrees of freedom tuning

```{r ss-tuning}
ss_cv <- kfold_rmse("ss", values = 4:10, data = Wage, K = 10)
colnames(ss_cv) <- c("df", "cv_rmse")
save_table(ss_cv, "tbl_ss_df_cv", digits = 3)

p_ss_cv <- ggplot(ss_cv, aes(df, cv_rmse)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 4:10) +
  labs(title = "Smoothing spline tuning by 10-fold CV",
       x = "Target effective df", y = "CV RMSE")
save_plot(p_ss_cv, "fig_ss_df_cv")

ss_cv
```

## Cross-validated and fixed-df smoothing splines

```{r ss-fit}
fit_ss_cv  <- smooth.spline(Wage$age, Wage$wage, cv = TRUE)
fit_ss_df6 <- smooth.spline(Wage$age, Wage$wage, df = 6)
ss_df <- age_grid
ss_df$cv_fit  <- predict(fit_ss_cv, x = age_grid$age)$y
ss_df$df6_fit <- predict(fit_ss_df6, x = age_grid$age)$y

p_ss_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.12) +
  geom_line(data = ss_df,
            aes(x = age, y = cv_fit, linetype = "CV chosen"),
            inherit.aes = FALSE,
            linewidth = 1) +
  geom_line(data = ss_df,
            aes(x = age, y = df6_fit, linetype = "df = 6"),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Smoothing spline fits", x = "Age", y = "Wage", linetype = "Fit")
save_plot(p_ss_fit, "fig_ss_fit")
p_ss_fit
```

```{r ss-summary-table}
ss_tbl <- data.frame(
  fit = c("CV chosen", "Fixed df = 6"),
  effective_df = c(fit_ss_cv$df, fit_ss_df6$df),
  spar = c(fit_ss_cv$spar, fit_ss_df6$spar)
)
save_table(ss_tbl, "tbl_ss_fit_summary", digits = 3)
kable(ss_tbl, digits = 3)
```

# Local regression

## Span tuning

```{r loess-tuning}
loess_cv <- kfold_rmse("loess", values = c(0.20, 0.35, 0.50, 0.65, 0.80), data = Wage, K = 10)
colnames(loess_cv) <- c("span", "cv_rmse")
save_table(loess_cv, "tbl_loess_span_cv", digits = 3)

p_loess_cv <- ggplot(loess_cv, aes(span, cv_rmse)) +
  geom_line() +
  geom_point() +
  labs(title = "LOESS span tuning by 10-fold CV",
       x = "Span", y = "CV RMSE")
save_plot(p_loess_cv, "fig_loess_span_cv")

loess_cv
```

## LOESS fits with different spans

```{r loess-fit}
fit_loess_02 <- loess(wage ~ age, data = Wage, span = 0.20, degree = 1)
fit_loess_05 <- loess(wage ~ age, data = Wage, span = 0.50, degree = 1)
loess_df <- age_grid
loess_df$fit_02 <- predict_loess_safe(fit_loess_02, age_grid, train_x = Wage$age)
loess_df$fit_05 <- predict_loess_safe(fit_loess_05, age_grid, train_x = Wage$age)

p_loess_fit <- ggplot(Wage, aes(age, wage)) +
  geom_point(alpha = 0.12) +
  geom_line(data = loess_df,
            aes(x = age, y = fit_02, linetype = "span = 0.20"),
            inherit.aes = FALSE,
            linewidth = 1) +
  geom_line(data = loess_df,
            aes(x = age, y = fit_05, linetype = "span = 0.50"),
            inherit.aes = FALSE,
            linewidth = 1) +
  labs(title = "Local regression with different spans", x = "Age", y = "Wage", linetype = "LOESS fit")
save_plot(p_loess_fit, "fig_loess_fit")
p_loess_fit
```

# Generalized additive models

## Variable selection and functional form selection

```{r gam-model-selection}
fit_lm_mult <- lm(wage ~ age + year + education, data = train)
fit_poly_mult <- lm(wage ~ poly(age, 4) + year + education, data = train)
fit_ns_mult <- lm(wage ~ ns(age, df = 5) + ns(year, df = 4) + education, data = train)
fit_gam_mult <- gam(wage ~ s(age, k = 6) + s(year, k = 5) + education,
                    data = train, method = "REML")

multivariable_compare <- data.frame(
  model = c("Linear age + year + education",
            "Polynomial age + year + education",
            "Natural splines for age and year",
            "GAM with smooth age and year"),
  aic = c(AIC(fit_lm_mult), AIC(fit_poly_mult), AIC(fit_ns_mult), AIC(fit_gam_mult)),
  test_rmse = c(
    rmse(test$wage, predict(fit_lm_mult, newdata = test)),
    rmse(test$wage, predict(fit_poly_mult, newdata = test)),
    rmse(test$wage, predict(fit_ns_mult, newdata = test)),
    rmse(test$wage, predict(fit_gam_mult, newdata = test))
  ),
  test_mae = c(
    mae(test$wage, predict(fit_lm_mult, newdata = test)),
    mae(test$wage, predict(fit_poly_mult, newdata = test)),
    mae(test$wage, predict(fit_ns_mult, newdata = test)),
    mae(test$wage, predict(fit_gam_mult, newdata = test))
  ),
  test_r2 = c(
    rsq(test$wage, predict(fit_lm_mult, newdata = test)),
    rsq(test$wage, predict(fit_poly_mult, newdata = test)),
    rsq(test$wage, predict(fit_ns_mult, newdata = test)),
    rsq(test$wage, predict(fit_gam_mult, newdata = test))
  )
)
save_table(multivariable_compare, "tbl_multivariable_model_compare", digits = 3)
kable(multivariable_compare, digits = 3)
```

## GAM term summary

```{r gam-term-summary}
fit_gam <- gam(wage ~ s(age, k = 6) + s(year, k = 5) + education,
               data = Wage, method = "REML")

gam_s_tbl <- as.data.frame(summary(fit_gam)$s.table)
gam_s_tbl$term <- rownames(gam_s_tbl)
rownames(gam_s_tbl) <- NULL
gam_s_tbl <- gam_s_tbl[, c("term", "edf", "Ref.df", "F", "p-value")]
colnames(gam_s_tbl) <- c("term", "edf", "ref_df", "F_stat", "p_value")
save_table(gam_s_tbl, "tbl_gam_term_summary", digits = 4)
kable(gam_s_tbl, digits = 4)
```

## GAM prediction profile and smooth term figure

```{r gam-age-profile}
age_profile$fit <- predict(fit_gam, newdata = age_profile)

p_gam_age <- ggplot(age_profile, aes(age, fit)) +
  geom_line(linewidth = 1) +
  labs(title = "GAM prediction as age varies",
       subtitle = paste("Year fixed at", round(median(Wage$year)), "and education fixed at", levels(Wage$education)[1]),
       x = "Age", y = "Predicted wage")
save_plot(p_gam_age, "fig_gam_age_profile")
p_gam_age
```

```{r gam-smooth-plots, fig.cap="GAM partial effect plots.", eval=TRUE}
pdf(file.path(fig_dir, "fig_gam_smooth_terms.pdf"), width = 8, height = 4.5)
par(mfrow = c(1, 2))
plot(fit_gam, select = 1, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
plot(fit_gam, select = 2, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
dev.off()

png(file.path(fig_dir, "fig_gam_smooth_terms.png"), width = 8, height = 4.5, units = "in", res = 300)
par(mfrow = c(1, 2))
plot(fit_gam, select = 1, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
plot(fit_gam, select = 2, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
dev.off()

par(mfrow = c(1, 2))
plot(fit_gam, select = 1, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
plot(fit_gam, select = 2, residuals = TRUE, pch = 16, cex = 0.35, shade = TRUE)
par(mfrow = c(1, 1))
```

## Logistic GAM and classification evaluation

```{r gam-logit}
fit_gam_logit <- gam(high_wage ~ s(age, k = 6) + s(year, k = 5) + education,
                     data = train, family = binomial, method = "REML")

logit_profile <- age_profile
logit_profile$prob <- predict(fit_gam_logit, newdata = age_profile, type = "response")

p_gam_logit <- ggplot(logit_profile, aes(age, prob)) +
  geom_line(linewidth = 1) +
  labs(title = "Logistic GAM: probability of high wage",
       subtitle = paste("Year fixed at", round(median(Wage$year)), "and education fixed at", levels(Wage$education)[1]),
       x = "Age", y = "Probability")
save_plot(p_gam_logit, "fig_gam_logit")
p_gam_logit
```

```{r class-compare}
fit_glm_linear <- glm(high_wage ~ age + year + education, data = train, family = binomial)
fit_glm_poly   <- glm(high_wage ~ poly(age, 3) + year + education, data = train, family = binomial)
fit_gam_class  <- gam(high_wage ~ s(age, k = 6) + s(year, k = 5) + education,
                      data = train, family = binomial, method = "REML")

prob_linear <- predict(fit_glm_linear, newdata = test, type = "response")
prob_poly   <- predict(fit_glm_poly, newdata = test, type = "response")
prob_gam    <- predict(fit_gam_class, newdata = test, type = "response")

class_compare <- data.frame(
  model = c("Linear logistic", "Polynomial logistic", "Logistic GAM"),
  log_loss = c(logloss(test$high_wage, prob_linear),
               logloss(test$high_wage, prob_poly),
               logloss(test$high_wage, prob_gam)),
  brier = c(brier(test$high_wage, prob_linear),
            brier(test$high_wage, prob_poly),
            brier(test$high_wage, prob_gam)),
  accuracy = c(accuracy(test$high_wage, prob_linear),
               accuracy(test$high_wage, prob_poly),
               accuracy(test$high_wage, prob_gam))
)
save_table(class_compare, "tbl_classification_model_compare", digits = 4)
kable(class_compare, digits = 4)
```

# One-variable age smoother comparison

```{r age-only-compare}
# Train-age models
fit_poly_age <- lm(wage ~ poly(age, 4), data = train)
fit_step_age <- make_step_model(train, bins = 4)
fit_bs_age   <- lm(wage ~ bs(age, knots = c(30, 40, 50, 60)), data = train)
fit_ns_age   <- lm(wage ~ ns(age, df = 5), data = train)
fit_ss_age   <- smooth.spline(train$age, train$wage, cv = TRUE)
fit_loess_age <- loess(wage ~ age, data = train, span = 0.35, degree = 1)

age_only_compare <- data.frame(
  model = c("Polynomial (degree 4)", "Step (4 bins)", "Regression spline",
            "Natural spline", "Smoothing spline", "LOESS"),
  test_rmse = c(
    rmse(test$wage, predict(fit_poly_age, newdata = test)),
    rmse(test$wage, predict_step_model(fit_step_age, test)),
    rmse(test$wage, predict(fit_bs_age, newdata = test)),
    rmse(test$wage, predict(fit_ns_age, newdata = test)),
    rmse(test$wage, predict(fit_ss_age, x = test$age)$y),
    rmse(test$wage, predict_loess_safe(fit_loess_age, data.frame(age = test$age), train_x = train$age))
  ),
  test_mae = c(
    mae(test$wage, predict(fit_poly_age, newdata = test)),
    mae(test$wage, predict_step_model(fit_step_age, test)),
    mae(test$wage, predict(fit_bs_age, newdata = test)),
    mae(test$wage, predict(fit_ns_age, newdata = test)),
    mae(test$wage, predict(fit_ss_age, x = test$age)$y),
    mae(test$wage, predict_loess_safe(fit_loess_age, data.frame(age = test$age), train_x = train$age))
  ),
  test_r2 = c(
    rsq(test$wage, predict(fit_poly_age, newdata = test)),
    rsq(test$wage, predict_step_model(fit_step_age, test)),
    rsq(test$wage, predict(fit_bs_age, newdata = test)),
    rsq(test$wage, predict(fit_ns_age, newdata = test)),
    rsq(test$wage, predict(fit_ss_age, x = test$age)$y),
    rsq(test$wage, predict_loess_safe(fit_loess_age, data.frame(age = test$age), train_x = train$age))
  )
)
save_table(age_only_compare, "tbl_age_only_model_compare", digits = 3)
kable(age_only_compare, digits = 3)
```

# Model evaluation notes

The main lesson is to separate three decisions:

1. **Variable selection**: which predictors enter the model?
2. **Functional-form selection**: should a predictor enter linearly, as a polynomial, as a step function, or as a smooth term?
3. **Complexity tuning**: if the predictor is modeled nonlinearly, how much flexibility should we allow?

A practical workflow is:

1. start with a scientific question and a basic plot,
2. choose candidate predictors,
3. decide which predictors deserve nonlinear treatment,
4. tune complexity with cross-validation or a validation set,
5. compare out-of-sample metrics,
6. inspect residuals, boundary behavior, and stability of conclusions.

# Session information

```{r session-info}
sessionInfo()
```
